Variance reduction for Markov chains via martingale representations

### 2020-05-02 aap/Revision/vr-mcmc-revision-new-new.tex
update

Under assumptions {\bf (H1)} and {\bf (H2)} the Markov chain $(X^x_{p})_{p \in \nset_0}$ is $V-$geometrically ergodic with $V(x) = 1+\|x\|^2$ and $\rho = \rme^{-\kappa\gamma}$ with $\kappa$ specified in Lemma~\ref{lem:v_ergodicity}. Due to lemma~\ref{lem:covariance},
\[
\PVar[\pi_n^N(f)]\lesssim \frac{1}{n(1-\rho^{1/2})}
\]
and the result of lemma follows from Taylor expansion of denominator.

@@

\begin{lem}
\label{lem:covariance}
Let $(X_p)_{p \in \nset_0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and $f(x)$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$. Let $\tilde{f}(x) = f(x) - \pi(f)$. Then
\begin{equation}
\label{eq:cov_v_ergodic}
\PCov_x \left[f(X_k),f(X_{k+s})\right] \leq C^2V^2(x)\rho^{2k+s} + 2C^2\rho^{k+s/2}\|\tilde{f}\|^2_{V^{\frac{1}{2}}}V(x) + 2C\rho^{s/2}\|\tilde{f}\|^2_{V^{\frac{1}{2}}}\pi(V)
\end{equation}
with constants $C > 0$ and $0 < \rho < 1$ from \eqref{eq:v_ergodicity}.
\end{lem}
\begin{proof}
Note that
\[
\PCov_x \left[f(X_k),f(X_{k+s})\right] = \PE_x\left[\tilde{f}(X_{k})\tilde{f}(X_{k+s})\right] + \left[\pi(f) - \PE_xf(X_{k+s})\right]\left[\PE_xf(X_k) - \pi(f)\right]
\]
Due to $V-$ergodicity \eqref{eq:v_ergodicity},
\[
\left|\left[\pi(f) - \PE_xf(X_{k+s})\right]\left[\PE_xf(X_k) - \pi(f)\right]\right| \leq C^2V^2(x)\rho^{2k+s}
\]
To bound the first term note that
\begin{multline*}
\left|\PE_x\left[\tilde{f}(X_{k})\tilde{f}(X_{k+s})\right] - \PCov_{\pi}\left[f(X_k),f(X_{k+s})\right]\right| = \left|\PE_x\left[\tilde{f}(X_{k})\tilde{f}(X_{k+s})\right] - \PE_\pi\left[\tilde{f}(X_{k})\tilde{f}(X_{k+s})\right]\right| \leq
\\
\leq \int\limits_{\rset^d}\left|\PE_y\left[\tilde{f}(X_0)\tilde{f}(X_{s})\right]\right| |\P^k(x,dy)-\pi(dy)|
\leq \{\eqref{eq:fp_cov}\}\leq 2C\rho^{s/2}\|\tilde{f}\|^2_{V^{\frac{1}{2}}}\int\limits_{\rset^d} V(y) |\P^k(x,dy)-\pi(dy)| \leq 
\\
\leq 2C^2\rho^{s/2}\|\tilde{f}\|^2_{V^{\frac{1}{2}}}\|\P^k(x,dy)-\pi(dy)\|_{V} \leq 2C^2\rho^{k+s/2}\|\tilde{f}\|^2_{V^{\frac{1}{2}}}V(x)
\end{multline*}
where $\left|\PCov_{\pi}\left[f(X_k),f(X_{k+s})\right]\right|$ can be bounded using \eqref{eq:fp_cov_stationary}. 
\end{proof}
Now we state and prove the main result of this section on the variance bound for the estimator $\pi_N^n(f)$ for $V-$geometrically ergodic Markov chain.
\begin{lem}
Let $(X_p)_{p \in \nset_0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and $f(x)$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$. Let $\tilde{f}(x) = f(x) - \pi(f)$ and assume that $n(1-\rho^{1/2}) > 1$. Then
\begin{equation}
\label{eq:vanilla_var}
\PVar_x \left[ \pi_{N}^n(f) \right] \leq \frac{C_1\pi(V)\|\tilde{f}\|^2_{V^{1/2}}}{n(1-\rho^{1/2})} + \mathcal{O}(\frac{1}{n^2(1-\rho^{1/2})})
\end{equation}
with constant $C_1 > 0$ and $0 < \rho <1$ from \eqref{eq:v_ergodicity}.
\end{lem}

>>>>

Under assumptions {\bf (H1)} and {\bf (H2)} the Markov chain $(X^x_{p})_{p \in \nset_0}$ is $V-$geometrically ergodic \eqref{eq:v_ergodicity} with $V(x) = 1+\|x\|^2$ and $\rho$ specified in Lemma~\ref{lem:v_ergodicity}. Due to lemma~\ref{lem:var_bound},
\[
\PVar[\pi_n^N(f)]\lesssim \frac{1}{n(1-\rho^{1/2})}
\]
and the result of lemma follows from Taylor expansion of denominator.

@@

\begin{lem}
\label{lem:covariance}
Let $(X_p)_{p \in \nset_0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and $f(x)$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$. Let $\tilde{f}(x) = f(x) - \pi(f)$. Then
\begin{equation}
\label{eq:cov_v_ergodic}
\PCov_x \left[f(X_k),f(X_{k+s})\right] \leq C^2V^2(x)\rho^{2k+s} + 2C^2\rho^{k+s/2}\|\tilde{f}\|^2_{V^{\frac{1}{2}}}V(x) + 2C\rho^{s/2}\|\tilde{f}\|^2_{V^{\frac{1}{2}}}\pi(V)
\end{equation}
with constants $C > 0$ and $0 < \rho < 1$ from \eqref{eq:v_ergodicity}.
\end{lem}
\begin{proof}
Note that
\[
\PCov_x \left[f(X_k),f(X_{k+s})\right] = \PE_x\left[\tilde{f}(X_{k})\tilde{f}(X_{k+s})\right] + \left[\pi(f) - \PE_xf(X_{k+s})\right]\left[\PE_xf(X_k) - \pi(f)\right]
\]
Due to $V-$ergodicity \eqref{eq:v_ergodicity},
\[
\left|\left[\pi(f) - \PE_xf(X_{k+s})\right]\left[\PE_xf(X_k) - \pi(f)\right]\right| \leq C^2V^2(x)\rho^{2k+s}
\]
To bound the first term note that
\begin{multline*}
\left|\PE_x\left[\tilde{f}(X_{k})\tilde{f}(X_{k+s})\right] - \PCov_{\pi}\left[f(X_k),f(X_{k+s})\right]\right| = \left|\PE_x\left[\tilde{f}(X_{k})\tilde{f}(X_{k+s})\right] - \PE_\pi\left[\tilde{f}(X_{k})\tilde{f}(X_{k+s})\right]\right| \leq
\\
\leq \int\limits_{\rset^d}\left|\PE_y\left[\tilde{f}(X_0)\tilde{f}(X_{s})\right]\right| |\P^k(x,dy)-\pi(dy)|
\leq \{\eqref{eq:fp_cov}\}\leq 2C\rho^{s/2}\|\tilde{f}\|^2_{V^{\frac{1}{2}}}\int\limits_{\rset^d} V(y) |\P^k(x,dy)-\pi(dy)| \leq 
\\
\leq 2C^2\rho^{s/2}\|\tilde{f}\|^2_{V^{\frac{1}{2}}}\|\P^k(x,dy)-\pi(dy)\|_{V} \leq 2C^2\rho^{k+s/2}\|\tilde{f}\|^2_{V^{\frac{1}{2}}}V(x)
\end{multline*}
where $\left|\PCov_{\pi}\left[f(X_k),f(X_{k+s})\right]\right|$ can be bounded using \eqref{eq:fp_cov_stationary}. 
\end{proof}
Now we state and prove the main result of this section on the variance bound for the estimator $\pi_N^n(f)$ for $V-$geometrically ergodic Markov chain.
\begin{lem}
\label{lem:var_bound}
Let $(X_p)_{p \in \nset_0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and $f(x)$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$. Let $\tilde{f}(x) = f(x) - \pi(f)$ and assume that $n(1-\rho^{1/2}) > 1$. Then
\begin{equation}
\label{eq:vanilla_var}
\PVar_x \left[ \pi_{N}^n(f) \right] \leq \frac{C_1\pi(V)\|\tilde{f}\|^2_{V^{1/2}}}{n(1-\rho^{1/2})} + \mathcal{O}(\frac{1}{n^2(1-\rho^{1/2})})
\end{equation}
with constant $C_1 > 0$ and $0 < \rho <1$ from \eqref{eq:v_ergodicity}.
\end{lem}
